----
title: "Steps Activity Analyses" 
author: "Jordan A. Kempker, MD, MSc " 
date: " r Sys.Date() "
----

#Introduction

This analysis uses data collected from a personal activity monitoring device.  Data includes the number of steps from a single individual collected in 5 minute intervals throughout each 24 hour day from October through November, 2012.




The variables in the dataset are:


* __steps:__ integer variable of steps for each 5-minute interval (missing values coded as __NA__)
* __date:__ character variable of date which measurement was recorded in YYY-MM-DD format
* __interval:__ integer variable of the time interval measurement was recorded in 2400 military time format



The dataset is stored in a comma-separated-value (CSV) with a total of 17,568 variables.



The R packages you will need to manually install (with ```install.packages()```) to run this analysis are:


* __stringr:__  for the ```str_pad()``` function
* __dplyr:__ for some nifty dataset management functions 
* __ggplot2:__ my favorite plotting system for R

Here is the code you will need to use:
```{r, eval = FALSE}
install.packages("stringr")

install.packages("dplyr")

install.packages("ggplot2")

```

```{r, include = FALSE}
library(stringr)
library(dplyr)
library(ggplot2)
```

---------------

#Downloading and Processing the Data

1. The data can be downloaded from my GitHub site, [here](https://github.com/jkempker/Activity/blob/master/repdata_data_activity.zip).

2. Set your own working directory with ```setwd()```.

3. Unzip the file and read into an R dataframe we will name __act__:

```{r, echo=TRUE}
unzip("./repdata_data_activity.zip")  
act <- read.csv("./activity.csv", stringsAsFactors = FALSE)
```

4. To join the date and time interval variables into one POSIXct formatted time variable
    + __interval__ is converted from integer to character and left padded with zeroes for fixed length = 4
    + __interval__ character is pasted to the __date__ then ```strptime()``` used to format

```{r, echo=TRUE}
act$interval_char <- as.character(act$interval)
act$interval_char <- str_pad(act$interval_char, 4, side = c("left"), pad ="0")
act$time_char <- paste(act$date, act$interval_char)
act$time <- as.POSIXct(strptime(act$time_char, format = "%Y-%m-%d %H%M"))
```

5. There is a lot of missingness for __steps__, so we will create __steps_imputed__ that takes the imputes the average amount of steps for each time interval, averaged over the entire period wherever __steps__ is missing.

```{r, echo=TRUE}
act$interval_char <- as.factor(act$interval_char)
means <- as.numeric(with(act, tapply(steps, INDEX=interval_char, mean, na.rm=TRUE)))

#few data steps to merge these means back to full set, matching by time interval factor level
interval_char<- levels(act$interval_char)
n <- data.frame(interval_char,means)
act <- merge(act, n, by = "interval_char")
act <- mutate(act, steps_imputed = ifelse(is.na(steps), means, steps))
```

6. For the below analysis, we will visually compare step activity of weekdays with weekends
    + Create a factor variable for day of the week using nifty ```weekdays()``` function
    + Use ```dplyr:mutate()``` again to create the __day_cat__ factor variable

```{r, echo=TRUE}
act <- mutate(act, day = weekdays(time))
act <- mutate(act, day_cat = ifelse(day=="Sunday", "weekend",
                                    ifelse(day == "Saturday", "weekend","weekday")))
act$day_cat <- as.factor(act$day_cat)
with(act, table(day, day_cat)) #verify this tricky ifelse mutate worked
```

7. Finally, clean up into a tidy dataframe, __activity__, and remove other temporary objects

```{r, echo=TRUE}
activity <- select(act, date, interval, time, steps, steps_imputed, day_cat) 
remove(act, n)
```

-------------
#Analysis
##Here we will pose a seris of questions and their answers.


1. Calculate the total number of steps taken per day.
    +Use ```aggregate()``` to sum up __steps__ over each __date__ (coerced into factor)
    +Other options would be to use ```tapply()``` or ```lapply(split())```.

```{r, echo=TRUE}
steps_per_day <- aggregate(steps ~ date, data = activity, FUN = sum, na.action = na.pass)
print(steps_per_day)
```

2. Make a histogram of total number of steps taken each day

```{r, echo=TRUE}
g <- ggplot(steps_per_day, aes(steps))+
    geom_histogram()
print(g)
```

3. Calculate the mean and median steps per day

```{r, echo=TRUE}
mean(steps_per_day$steps, na.rm = TRUE)
median(steps_per_day$steps, na.rm = TRUE)
```

4. Make a time series plot with 5-minute interval on x-axis and average steps (averaged per interval over entire study period) on y-axis

```{r, echo=TRUE}
avg_steps_int <- aggregate(steps ~ interval, data = activity, FUN = mean)
ggplot(avg_steps_int, aes(interval, steps))+
    geom_path()+
    labs(x = "5-minute time interval", y = "Average steps taken")
```

5. Which interval has the most steps on average?

```{r, echo=TRUE}
with(avg_steps_int, max(steps), na.rm = TRUE)
filter(avg_steps_int, steps == max(steps))
```

6. Report the total number of missing values for steps
    +Here the missings are summed up as __TRUEs__ with the ```is.na()``` function

```{r, echo=TRUE}
table(is.na(activity$steps))
```

7. Using the __steps_imputed__ variable created above, remake a histogram of step counts per interval and calculate __mean__ and __median__

```{r, echo=TRUE}
g <- ggplot(activity, aes(steps_imputed))+
    geom_histogram()
print(g)
mean(activity$steps_imputed)
median(activity$steps_imputed)
```

8. Using __day_cat__ variable created above, which creates a factor varialbe grouping days by weeekends and weekdays, create a multipanel time series plot of average steps per time interval (averaged over entire study period grouped by weekday vs. weekend) to compare.

```{r, echo=TRUE}
avg_steps_weekend <- aggregate(steps ~ day_cat + interval, data = activity, FUN = mean)

g <- ggplot(avg_steps_weekend, aes(interval, steps))+
    geom_path()+
    facet_grid( day_cat ~ .)
print(g)
```
